Load required library

```{r}
packages_to_install=c("plotly", "circular")
install.packages(setdiff(packages_to_install, rownames(installed.packages())))
```

Load the libraries
```{r}
require(plotly)
require(circular)
```
 ---------------------------------------- #
Import data and convert spikes to data matrix
 ---------------------------------------- #

Load the data
```{r}
load(file= "ephysData.Rdata")
```

spike count for all trials
note about sizes:
data$EVENTS is [ neurons gratings trials ]

extract sizes of the data matrix

```{r}
nNeurons <- dim(DATA$EVENTS)[1]
nGratings <- dim(DATA$EVENTS)[2]
nTrials <- dim(DATA$EVENTS)[3]
```

initialize
```{r}
totalSpikeCount <- array(0, dim = c(nNeurons, nGratings, nTrials))
```

loop through all elements of the data matrix
```{r}
for (neuroni in 1:nNeurons) {
  for (grati in 1:nGratings) {
    for (triali in 1:nTrials) {
      # count the number of spikes in each cell
      totalSpikeCount[neuroni, grati, triali] <- length(DATA$EVENTS[[neuroni, grati, triali]])
    }
  }
}
```

and now using lapply instead of for loop
```{r}
totalSpikeCount2 <- array(unlist(lapply(DATA$EVENTS, length)),
                          dim = c(nNeurons, nGratings, nTrials))
```

compare them to show they're equal
```{r}
dim(totalSpikeCount)
dim(totalSpikeCount2)
differenceMatrix <- totalSpikeCount - totalSpikeCount2
sum(differenceMatrix)
```

------------------------------------------ #
Histograms of spike counts       #
------------------------------------------ #


```{r}

hist(totalSpikeCount, breaks = 40,
     main = "Including zero-spike trials",
     xlab = "Number of spikes",
     ylab = "Count")
```


Exclude trials with no spikes
```{r}
hist(totalSpikeCount2[totalSpikeCount != 0], breaks = 40,
     main = "Excluding zero-spike trials",
     xlab = "Number of spikes",
     ylab = "Count")
```

------------------------------------------ #
Tuning curve                     #
------------------------------------------ #


pick a unit and compute its tuning curve
pick a neuron at random
```{r}
randomunit <- sample(1:nNeurons, 1)

# compute the average spike count over all trials per orientation
averageSpikes <- apply(totalSpikeCount2[randomunit,,], c(1,2), mean)

# find the maximum response
maxresp <- which.max(averageSpikes)

# visualize
gradientOrient <- seq(0, 330, by = 30)

# bar plot
barplot(averageSpikes, names.arg = gradientOrient,
        xlab = "Gradient orientation",
        ylab = "Average spike count",
        main = paste("Unit number", randomunit,
                     "\"prefers\"", gradientOrient[maxresp], "degrees"),
        col = "blue")

```

Generate a polar plot
# https://plotly.com/r/polar-chart/

```{r}
theta <- circular(deg2rad(gradientOrient + 90))
r <- circular(averageSpikes)
plot(theta, r,
     type = "p", pch = 16, col = "blue",
     main = paste("Unit number", randomunit,
                  "\"prefers\"", gradientOrient[maxresp], "degrees"),
     ylab = "Average spike count")
```
------------------------------------------ #
Visualize a spatial map          #
------------------------------------------ #

gather the data
first map is total number of spikes
then find all the unique units in the dataset

```{r}
spikesPerChan <- rowMeans(totalSpikeCount2, na.rm = TRUE)

uniqueunits <- unique(data$CHANNELS[,1])
```

create the data matrix
```{r}
spikesMap <- matrix(0, nrow = nrow(data$MAP), ncol = ncol(data$MAP))

for (uniti in 1:length(uniqueunits)) {
  whichchans <- data$CHANNELS[,1] == uniqueunits[uniti]
  indices <- which(data$MAP == uniqueunits[uniti], arr.ind = TRUE)
  spikesMap[indices] <- mean(spikesPerChan[whichchans], na.rm = TRUE)
}

# visualize
image(spikesMap, col = heat.colors(256),
      xlab = "Column", ylab = "Row",
      main = "Firing rate map", axes = FALSE)
axis(1, at = seq(1, ncol(data$MAP), by = 10))
axis(2, at = seq(1, nrow(data$MAP), by = 10))
```
------------------------------------------ #
Visualize a spatial map          #
------------------------------------------ #

# compute the average spike count over all trials per orientation
```{r}
averageSpikes <- rowMeans(totalSpikeCount, dims = 2)

# find the maximum response and convert to degrees
maxresp <- apply(averageSpikes, 1, which.max)
maxresp <- gradientOrient[maxresp]

# create the data matrix
orientationMap <- matrix(NA, nrow = nrow(data$MAP), ncol = ncol(data$MAP))

for (uniti in 1:length(uniqueunits)) {
  whichchans <- data$CHANNELS[,1] == uniqueunits[uniti]
  indices <- which(data$MAP == uniqueunits[uniti], arr.ind = TRUE)
  orientationMap[indices] <- mean(maxresp[whichchans])
}

# visualize using a circular colormap
image(orientationMap, col = heat.colors(256),
      xlab = "Column", ylab = "Row",
      main = "Orientation map", axes = FALSE)
axis(1, at = seq(1, ncol(data$MAP), by = 10))
axis(2, at = seq(1, nrow(data$MAP), by = 10))
```

-------------------------------------------- #
Optional additional exercises
 -------------------------------------------- #

1) Filter out trials with fewer than 5 spikes. Does this qualitatively
   change the results? Try soft-coding to be able to specify any number
   of spikes to filter out.
 
2) Repeat the code for different datasets (monkeys 2-3). Put the results 
  into different figures to compare the findings across datasets. What results
  can be directly compared?

3) Import the stimulus images and visualize the first frame of each
   stimulus video.



```